---
title: "Lecture 5"
output: slidy_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```
#############
##  t-test

```{r, echo = TRUE}
daily.intake <- c(5260,5470,5640,6180,6390,6515, 6805,7515,7516,8230,8770)
mean(daily.intake)
sd(daily.intake)
quantile(daily.intake)
```
You might wish to investigate whether the women’s energy intake deviates
systematically from a recommended value of 7725. How to do it?

##  t-test

```{r , echo = TRUE}
t.test(daily.intake,mu=7725)
#t.test(daily.intake,mu=7725,alternative = "greater")
#t.test(daily.intake,mu=7725,alternative = "less")
```

## Wilcoxon signed-rank test
The one-sample Wilcoxon signed-rank test is a non-parametric statistical hypothesis test used to test the location of a set of samples $X_1, \dots, X_n$.
$$ T = \sum_{i=1}^n \mathrm{sgn}(X_i) R_i, $$
where $R_1, \dots, R_n$ are the ranks such that $0<|X_{R_1}|<|X_{R_2}|<\dotsb<|X_{R_n}|$

```{r,echo=TRUE}
wilcox.test(daily.intake, mu=7725)
```
If this test is unfamiliar to you, please read https://online.stat.psu.edu/stat414/node/319/.

## Two sample t-test 
```{r,echo=T}
x1 = rnorm(100,0,1)
x2 = rnorm(100,1,1)
x = c(x1,x2)
group = c(rep(1,100),rep(0,100))
t.test(x~group)
#t.test(x~group,var.equal=T)
```


## Comparing variances

```{r,echo=T}
var.test(x~group)
```

## The paired t test
Paired tests are used when there are two measurements on the same experimental
unit. The theory is essentially based on taking differences and
thus reducing the problem to that of a one-sample test
$$ t = \dfrac{\bar{X}_D - \mu_0}{s_D/\sqrt{n}}, $$
where $\bar{X}_D$ and $s_D$ are the average and standard deviation of the 
differences between all pairs.
```{r,echo=T}
t.test(x1, x2, paired=T)
```
It is virtually identical to
that of a one-sample t test on the elementwise differences.

## The matched-pairs Wilcoxon test

```{r,echo=T}
wilcox.test(x1, x2, paired=T)
```

## Simple linear regression
The simple linear regression model is 
$$y_i = \beta_0 + \beta_1 x_i +\epsilon_i, \epsilon_i \sim N(0,\sigma^2).$$
```{r,echo=T,ig.height=3, fig.width=6}
x = rnorm(50,0,1)
y = 1 + 2*x + rnorm(50,0,0.1)
```
## Simple linear regression

```{r,echo=T,ig.height=3, fig.width=6}
plot(x,y)
```

## Fit the regression model 
```{r,echo=T,ig.height=3, fig.width=6}
lm(y~x)
```

## Summary
```{r,echo=T,ig.height=3, fig.width=6}
fit = lm(y~x)
summary(fit)
```


## Residuals
```{r,echo=T,ig.height=3, fig.width=6}
plot(fit)
```

## Residual Check 
```{r,echo=T,ig.height=3, fig.width=6}
qqnorm(resid(fit))
```

## Correlation
```{r,echo=T,ig.height=3, fig.width=6}
cor(x1,x2)
cor.test(x1,x2)
```

## Spearman
This is obtained quite simply by replacing
the observations by their rank and computing the correlation.
```{r,echo=T,ig.height=3, fig.width=6}
cor.test(x1,x2,method = "spearman")
```







