---
title: "Tutorial 07: Presentations and Shiny Web App"
author: "<br>Department of Statistics and Actuarial Science</br> The University of Hong Kong"
output: ioslides_presentation
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, tidy = TRUE)
```



# Presentations from R Markdown


---

- Create a slide show broken up into sections by using the `#` and `##` heading tags.
- Create a new slide without a header using a horizontal rule (`---`).



## Ioslides Presentation

  [Reference](https://bookdown.org/yihui/rmarkdown/ioslides-presentation.html)

  - `f`: Enable fullscreen mode
  - `w`: Toggle widescreen mode
  - `o`: Enable overview mode
  - `h`: Enable code highlight mode
  - `p`: Show presenter notes  
  \
  
  Nice default appearance. Suitable for making presentations.



## Slidy Presentation

  [Reference](https://bookdown.org/yihui/rmarkdown/slidy-presentation.html)

  - `c`: Show table of contents
  - `f`: Toggles the display of the footer
  - `a`: Toggles display of current vs all slides (useful for printing handouts)
  - `s`: Make fonts smaller
  - `b`: Make fonts larger  
  \ 
  
  Suitable for making presentations to be shared or printed.




# Shiny Web App

## Empty Shiny App

```{r}
library(shiny)

# Define UI ----
ui <- fluidPage(
  
)

# Define server logic ----
server <- function(input, output) {
  
}

# Run the app ----
shinyApp(ui = ui, server = server)
```



## Layout

- `fluidPage` creates a display that automatically adjusts to the dimensions of user's browser window  
\ 

```{r}
ui <- fluidPage(
  titlePanel("Hello Shiny!")
)

shinyApp(ui = ui, server = function(input, output){})
```





---

### **Sidebar Layout**

- `sidebarLayout` 
  - `sidebarPanel`
  - `mainPanel`  
\ 

```{r}
ui <- fluidPage(

  titlePanel("Hello Shiny!"),

  sidebarLayout(
    
    position = "left", # by default

    sidebarPanel("sidebar panel"),

    mainPanel("main panel")
  )
)

shinyApp(ui = ui, server = function(input, output){})
```




---

### **Grid Layout**

- `fluidRow`
  - `column`
    - `wellPanel`  
\ 


```{r}
ui <- fluidPage(

  titlePanel("Hello Shiny!"),

  fluidRow(
  
    column(4,
      wellPanel(
        "well panel"
      )       
    ),

    column(8,
      "normal column"
    )
  ),
  
  fluidRow(
  
    column(12,
      wellPanel(
        "well panel in row 2"
      )       
    )
  )
)

shinyApp(ui = ui, server = function(input, output){})
```



---

### Other Layouts

[Official Layout Guide](https://shiny.rstudio.com/articles/layout-guide.html)



## Control Widgets

- `actionButton`	Action Button
- `checkboxGroupInput`	A group of check boxes
- `checkboxInput`	A single check box
- `dateInput`	A calendar to aid date selection
- `dateRangeInput`	A pair of calendars for selecting a date range
- `fileInput`	A file upload control wizard
- `helpText`	Help text that can be added to an input form
- `numericInput`	A field to enter numbers
- `radioButtons`	A set of radio buttons
- `selectInput`	A box with choices to select from
- `sliderInput`	A slider bar
- `submitButton`	A submit button
- `textInput`	A field to enter text


---

### First two arguments for each widget function

- a name for the widget
- a label appear with the widget in your app

---

```{r}
library(shiny)

# Define UI ----
ui <- fluidPage(
  titlePanel("Basic widgets"),
  
  fluidRow(
    
    column(3,
           h3("Buttons"),
           actionButton("action", "Action"),
           br(),
           br(), 
           submitButton("Submit")),
    
    column(3,
           h3("Single checkbox"),
           checkboxInput("checkbox", "Choice A", value = TRUE)),
    
    column(3, 
           checkboxGroupInput("checkGroup", 
                              h3("Checkbox group"), 
                              choices = list("Choice 1" = 1, 
                                             "Choice 2" = 2, 
                                             "Choice 3" = 3),
                              selected = 1)),
    
    column(3, 
           dateInput("date", 
                     h3("Date input"), 
                     value = "2014-01-01"))   
  ),
  
  fluidRow(
    
    column(3,
           dateRangeInput("dates", h3("Date range"))),
    
    column(3,
           fileInput("file", h3("File input"))),
    
    column(3, 
           h3("Help text"),
           helpText("Note: help text isn't a true widget,", 
                    "but it provides an easy way to add text to",
                    "accompany other widgets.")),
    
    column(3, 
           numericInput("num", 
                        h3("Numeric input"), 
                        value = 1))   
  ),
  
  fluidRow(
    
    column(3,
           radioButtons("radio", h3("Radio buttons"),
                        choices = list("Choice 1" = 1, "Choice 2" = 2,
                                       "Choice 3" = 3),selected = 1)),
    
    column(3,
           selectInput("select", h3("Select box"), 
                       choices = list("Choice 1" = 1, "Choice 2" = 2,
                                      "Choice 3" = 3), selected = 1)),
    
    column(3, 
           sliderInput("slider1", h3("Sliders"),
                       min = 0, max = 100, value = 50),
           sliderInput("slider2", "",
                       min = 0, max = 100, value = c(25, 75))
    ),
    
    column(3, 
           textInput("text", h3("Text input"), 
                     value = "Enter text..."))   
  )
  
)

# Define server logic ----
server <- function(input, output) {
  
}

# Run the app ----
shinyApp(ui = ui, server = server)
```




## Reactive Output

### **Output functions**

Turn R objects into output for your user interface

- `dataTableOutput`	DataTable
- `htmlOutput	raw` HTML
- `imageOutput`	image
- `plotOutput`	plot
- `tableOutput`	table
- `textOutput`	text
- `uiOutput`	raw HTML
- `verbatimTextOutput`	text


---

```{r}
fluidPage(

  # App title ----
  titlePanel("Hello Shiny!"),

  # Sidebar layout with input and output definitions ----
  sidebarLayout(

    # Sidebar panel for inputs ----
    sidebarPanel(

      # Input: Slider for the number of bins ----
      sliderInput(inputId = "bins",
                  label = "Number of bins:",
                  min = 1,
                  max = 50,
                  value = 30)

    ),

    # Main panel for displaying outputs ----
    mainPanel(

      # Output: Histogram ----
      plotOutput(outputId = "distPlot")

    )
  )
)
```

---

### **Render functions**

Each entry to `xxxOutput` functions should contain the output of one of Shiny’s `renderXxx` functions.

- `renderDataTable`	dataTable
- `renderImage`	images (saved as a link to a source file)
- `renderPlot`	plots
- `renderPrint`	any printed output
- `renderTable`	data frame, matrix, other table like structures
- `renderText`	character strings
- `renderUI`	a Shiny tag object or HTML

---

```{r}
ui <- fluidPage(

  # App title ----
  titlePanel("Hello Shiny!"),

  # Sidebar layout with input and output definitions ----
  sidebarLayout(

    # Sidebar panel for inputs ----
    sidebarPanel(

      # Input: Slider for the number of bins ----
      sliderInput(inputId = "bins",
                  label = "Number of bins:",
                  min = 1,
                  max = 50,
                  value = 30)

    ),

    # Main panel for displaying outputs ----
    mainPanel(

      # Output: Histogram ----
      plotOutput(outputId = "distPlot")

    )
  )
)

server <- function(input, output) {
  output$distPlot <- renderPlot({

    x    <- faithful$waiting
    bins <- seq(min(x), max(x), length.out = input$bins + 1)

    hist(x, breaks = bins, col = "#75AADB", border = "white",
         xlab = "Waiting time to next eruption (in mins)",
         main = "Histogram of waiting times")

    })

}

shinyApp(ui = ui, server = server)
```


## Exercise 

Change the slider widget in the last Shiny app (for histogram of waiting times) to a select boxe.